HiC Contact Map Comaprison Using Graphlet Approach

### 2018-05-23 papers/manuscripts/manuscript.tex
final draft of the manuscript

In this study, we investigated
dissimilarities between normal cells and cancerous cells,
through analyzing HiC contact maps. 
Our results show that certain orbit distributions
have significanly higher correlation 
between Leukemic cells.

@@

\section{Introduction}

Ideally, it is desirable to compare 3D structures of 
cell in order to make such comparisons.
However, the main challenge is that 
3D structure of a cell is not readily available. 
In orther to find dissimilarities in the 3D structure of 
chromosomes, we used HiC dataset.
The HiC method, which was developed by 
\cite{lieberman2009comprehensive}, captures interactions between 
chromosomal fragments in kilobase resolution. Based on HiC data, an
\textit{interaction frequency (IF) } matrix can be developed 
between \textit{loci} at a desired resolution.
A cell $IF_{ij}$ in an interaction frequency matrix captures 
the number of interaction detected
in HiC dataset between locus $i$ and locus $j$ in the genome.
An interaction matrix can be used to develop both 
inter- and intra-chromosomal interaction matrices.
We believe differences in interaction matrices can 
be found between normal cells and cancerous ones.

Recently, there has been many efforts at predicting the 3D conformation
of chromosomes using HiC data. 
\cite{noble2011three, rousseau2011three, hu2013bayesian, varoquaux2014statistical, trieu2014large, zhang20133d, lesne20143d, bau2011three, adhikari2016chromosome3d}.
These efforts, although different in
approach, usually translate 
interaction frequencies in contact maps as an inverse measure of 
distance. In particular \cite{adhikari2016chromosome3d} developed
a method, Chromosome3D which uses
distance geometry simulated annealing (DGSA) in order to
reconstruct genome structure. These reconstructed 3D 
structures could be used in order to compare normal and malignant
cell lines. However, the main obstacle to doing so is that the
process of reconstruction is computationally
expensive. Almost all reconstruction methods rely on either 
deterministic or heuristic optimization in order to find the best
confomation that matches contact maps. Also, there is no real 
way of verifying whether the reconstruction results are accurate.
Fluorescence in situ hybridizaiton
(FISH) is used for investigating 3D configuration of chromosomes.
\cite{gasser2002visualizing, lanctot2007dynamic, mateos2009spatially}
However, this method can only be used locally and cannot map
the whole structure of the chromosomes.

Graphlet comparison is a novel method used to compare 
large networks in order to
find local similarities in them.
Authors of \cite{prvzulj2007biological} 
provide a new measure of PPI
network comparison
based on 73 constraints. This is used in order to compare two large
networks in order to detect similarities.

@@

The authors of \cite{di2010fast} generalized the idea of graphlets to 
ordered graphs were the nodes are labeled in ascending order.
As can be viewed, there are a total of 14 orbits for graphlets of size
2 and 3 since the label of graphlets is also included in toplogy.
In the new definition, $d_v^i$ denotes the number of orbit $i$ touches 
node $v$. Each node, is then assigned a vector of length 14 
\footnote{number of orbits in graphlets of size 2 and 3}
$(d_v^1, d_v^2, ..., d_v^{14})$ 
and similarity of two nodes in two contact maps can be compared by
how geometrically close their corresponding vectors are.

\subsubsection{Notations}
In this paper, matrices and vectors are represented with bold
capital and bold small letters respectively.
matrix rows and columns are represented by a \textit{dot}
notation. For example, the $i$th row of matrix $M$ is
denoted by $M_{i.}$ and its $j$th column is represented
by $M_{.j}$.

We denote the set of all contact maps in cell line $T$ with 
$\mathbb{C}^T$. If no particular cell line is addressed, the
subscripts are dropped.
Any arbitrary member of $\mathbb{C}$ is denoted by 
$C_{ij}$, where $i$ and $j$ ($j \ge i$) represent the two chromosomes involved. 
In human cells this set contains a total of 276 contact maps,
23 of which are intra-chromosomal and the rest are inter-chromosomal.
For ease of representations, intra-chromosomal contact maps are
distinguished by a single superscript, so we have $C_{i,i} =
C_i$.

We denote the number of loci in a chromosome $i$ by $N_i$.
The set of all loci involved in contact map $C_{ij}$ is denoted 
by $\mathbb{V}_{ij}$.
In intra-chromosomal contact maps, $\mathbb{V}_{i,i}$ containts only the 
loci of that particular chromosome ($|\mathbb{V}_i| = N_i$), while in 
inter-chromosomal contact maps $\mathbb{V}_{ij}$ contains the loci in
the both of chromosomes involved ($|\mathbb{V}_{ij}| = N_i + N_j$).

@@

\section{Materials and Methods}
We re-used Leukemic Hi-C libraries 
created in \cite{wang2013properties}
These libraries we sequenced 
for cases of primary human B-acute
lymphoblastic leukemia (B-ALL or ALL), 
the MHH-CALL-4 B-ALL cell
line (CALL4), 
and the follicular lymphoma cell-line (RL).
Just as \cite{wang2013properties}, 
we used normal B-cell line (GM068990)
from \cite{lieberman2009comprehensive} 
for our comparisons.
We created contact maps of resolution
500kb and normalized it using 
the \texttt{iced} package in python
developed by \cite{servant2015hic}.
We also performed and extra normalization by
performing network deconvolution \cite{feizi2013network}
in order to remove
the effect of indirect interactions.

@@

Instead, in order to threshold the matrix so that both global and local patterns are
captured, we borrowed the concept of \textit{adaptive thresholding} from image 
processing context. In this method, in order to be set, a pixel should have
an intensity larger than the average of non-zero intensities in its
\textit{neighborhood}. The neighborhood is defined by a sliding kernel 
that passes through the contact map with the pixel at its middle at 
each step. Figure \ref{local_thresholded_chr1_chr1} demonstrates result of 
this thresholding approach for intra-chromosomal contact maps of chromosome 1.
Refer to supplementary material for all 23 interchromosomal thresholding
results.
\begin{figure}[t]
    \centering
    \includegraphics[width=.5\textwidth]{figures/local_thresholded_chr1_chr1.png}
    \caption{Result of thresholding interchromosomal contact map of chromosome 1
    using a kernels of size $5 \times 5$ for all cell lines. 
    The first row shows the thresholded
    maps. Second and third rows demonstrate pair-wise similarities and 
    differences between contact maps respectively.}
    \label{local_thresholded_chr1_chr1}
\end{figure}

\subsection{Orbit Extraction}
Once the thresholded contact maps are obtained, graphlets and orbits can be 
extracted. We used the \texttt{orca} package in \texttt{R} programming 
language to extract the graphlets. As a result of graphlet extraction, 
For each loci in each contact map, a \textit{signature vector} of size
73 is created. Thus for each cell line, we would have 276 
\textit{signature matrices} of
size $|V^{ij}|\times 73$, where $V^{ij}$ is the number of loci
involved in contact map between chromosomes $i$ and $j$. Figure
\ref{fig:graphlet_extraction} illustrates the process and results
of signature matrix extraction schematically.
\begin{figure}
    \centering
    \includegraphics[width=.5\textwidth]{figures/graphlet_extraction.png}
    \caption{Graphlet extraction for the four cell lines. For each
    loci in each contact map between chromosomes $i$ and $j$, 
    the signature vectors of length 73 are extracted, resulting 
    in a \textit{signature matrix} of size $|V^{ij}| \times 73$,
    where $V^{ij}$ 
    is the number of loci involved.}
    \label{fig:graphlet_extraction}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=.5\textwidth]{figures/graphlet_distance_schema.png}
    \caption{Calculating pair-wise loci distances. For each loci (row) 
    in each
    contact map in MIT cell line, its distance is calculated based on
    equation \ref{eq:distance_total} with the corresponding loci in
    leukemic cells. The result of this process is a 
    \textit{signature distance vector} of size
    $|V^{ij}| = N_i+N_j$ for each contact map.
    }
    \label{graphlet_distance_schema}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=.5\textwidth]{figures/graphlet_correlation_schema.png}
    \caption{Calulating pair-wise orbit correlations. For each orbit (column)
    in each contact map in MIT cell line, its correlation with the
    same orbit in the same contact map 
    in leukemic cells is calculated. The result of
    this process is a \textit{signature correlation} vector of size
    73 which captures how similar frequencies of two orbits are.
    In order to test our second hypothesis, we calculated averages
    across contact maps (along the vertical red arrow) to test 
    hypothesis \ref{eq:h2a}
    and across orbits
    (along the horizontal red arrows) to test hypothesis \ref{eq:h2b}.}
    \label{graphlet_correlation_schema}
\end{figure}
%$
%\begin{pmatrix}
%    {S}^A_{\scriptscriptstyle i,j,1,1}         &
%    \cdots                                     &    
%    {S}^A_{\scriptscriptstyle i,j,1,72}        \\
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    \vdots                                     &                
%                                               &      
%    \vdots                                     \\
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   {S}^A_{\scriptscriptstyle i,j,N_1,1}    &
%   \ddots                                            &    
%   {S}^A_{\scriptscriptstyle i,j,N_1, 72} \\
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    \vdots                                     &                
%                                               &      
%    \vdots                                     \\
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   {S}^A_{\scriptscriptstyle i,j,N_1+N_2,1}   &
%    \cdots                                &
%   {S}^A_{\scriptscriptstyle i,j,N_1+N_2, 72}
%\end{pmatrix}
%$
%
%$
%\begin{pmatrix}
%    {C}^A_{\scriptscriptstyle i,j,1,1}     &
%    \cdots    &
%    {C}^A_{\scriptscriptstyle i,j,1,N_1} \\
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    \vdots                             &
%    \ddots&
%    \vdots                             \\
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    {C}^A_{\scriptscriptstyle i,j,N_1,1}   &
%    \cdots&                  
%    S^A_{\scriptscriptstyle i,j,N_1, N_1}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\end{pmatrix}
%$

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{.5\textwidth}
        \includegraphics[width=\textwidth]{figures/orbit-distances_chr14.png}
    \end{subfigure}
    \caption{   
     }
    \label{fig:orbit_distances_chr14}
\end{figure}

@@

We performed one-way ANOVA statistical test to 
see if there are significant 
difference between cancer-normal and cancer-cancer pairs.
We found that the difference statistically significant difference.
($F(1, 1654) = 20.49, p < 0.0001$).
As illustrated in figure \ref{fig:orbit_distances_all_a},
we can see that normal-cancer cells have higher distance
from each other than cancer-cancer pairs.

@@

Figure \ref{fig:orbits_correlations} demonstrates that certain
orbits of Leukemic have higher correlation to each other 
than to the
normal MIT cell. In fact our statistical analysis shows that 
\textit {for orbits NOT in $\mathbb{Q}$, 
intra-leukemic orbit correlations are significantly higher
than leukemic-normal orbit correlations}. This implies
there are significant differences between normal and
leukemic cells in terms of their local structure.

>>>>

In this study, we investigated
dissimilarities between normal cells and leukemic cells
in terms of their three dimensional conformation.
We first thresholed HiC data corresponding to
one normal cell line and four leukemic cells lines.
We then used the thresholded data to extract the 
first 72 orbits for all loci in each contact map.
We then measured pairwise graphlet distances as
well as pairwise graphlet distribution correlations
for each pair of cells and compared them using 
statistical methods.
Our results show that normal-cancer pairs have
significantly higher dinstances from each other
than cancer-cancer pairs.
Also, we concluded that for certain orbits,
cancer-cancer pairs demonstrate higher correlations
than normal-cancer pairs in terms of orbit frequency
distributions.

@@

\section{Introduction}
This reasearch is an effort to compare the three-dimensional
structure of cell genomes. To the best of our knowledge
little research has dealt with quantitative comparison
of spatial conformation of chromosomes.
Ideally, it is desirable to compare 3D structures of 
cell in order to make such comparisons.
However, the main challenge in this regard
is that 
3D structure of a cell is not readily available. 
In order to find dissimilarities in the 3D structure of 
chromosomes, we used HiC dataset.
The HiC method, which was developed by 
\cite{lieberman2009comprehensive}, captures interactions between 
chromosomal fragments in kilobase resolution. 
Based on HiC data, a \textit{contact map} or
\textit{interaction frequency (IF) } matrix can be developed 
between \textit{loci} at a desired resolution.
A cell $IF_{ij}$ in an interaction frequency matrix captures 
the number of interaction detected
in HiC dataset between locus $i$ and locus $j$ in the genome.
An interaction matrix can be used to develop both 
inter- and intra-chromosomal interaction matrices.
We have found statistically significant
differences between interaction matrices of
normal cells and cancerous ones.

Recently, HiC data has been mostyly used
in efforts to predict the 3D conformation
of chromosomes using HiC data. 
\cite{noble2011three, 
rousseau2011three, 
hu2013bayesian, 
varoquaux2014statistical, 
trieu2014large, 
zhang20133d, 
lesne20143d, 
bau2011three, 
adhikari2016chromosome3d}.
These efforts, although different in
approach, usually translate 
interaction frequencies in contact maps as an inverse measure of 
distance. For example,  \cite{adhikari2016chromosome3d} developed
a method, Chromosome3D which uses
distance geometry simulated annealing (DGSA) in order to
reconstruct genome structure. These reconstructed 3D 
structures could be used in order to compare normal and malignant
cell lines. However, the main obstacle to doing so is that the
process of reconstruction is computationally
expensive. Almost all reconstruction methods rely on either 
deterministic or heuristic optimization approaches
in order to find the best
confomation that matches contact maps. Also, there is no real 
way of verifying whether the reconstruction results are accurate.
The only method of verifying such results is
Fluorescence in situ hybridizaiton
(FISH) which can capture the actual
3D configuration of chromosomes.
\cite{gasser2002visualizing, lanctot2007dynamic, mateos2009spatially}
However, this method can only be used locally and cannot map
the whole structure of the chromosomes.

Recently, graphlet comparison has emerged
as a novel method for comparing
large networks in order to
find local similarities in them.
Authors of \cite{prvzulj2007biological} 
provide a new measure of protein-protein
interaction (PPI)
network comparison
based on 73 constraints. The method
is used in order to compare two large
PPI networks in order to detect similarities.

@@

In this paper, we applied the same approach of graphlet distance
and graphlet distribution comparison on HiC data
in order investigate genome spatial confomations differences
between normal and leukemic cells.

@@

\subsubsection{Notations}
In this paper, matrices and vectors are represented using bold
capital and bold small letters respectively.
Matrix rows and columns are represented by a \textit{dot}
notation. For example, the $i$th row of matrix $M$ is
denoted by $M_{i.}$ and its $j$th column is represented
by $M_{.j}$.

We denote the set of all contact maps in cell line $K$ with 
$\mathbb{C}^K$. If no particular cell line is addressed, the
subscripts are dropped.
Any arbitrary member of $\mathbb{C}$ is denoted by 
$\mathbf{C}_{ij}$, where $i$ and $j$ ($j \ge i$) 
represent the two chromosomes involved. 
In human cells this set contains a total of 276 contact maps,
23 of which are intra-chromosomal and the rest are inter-chromosomal.
For ease of representations, intra-chromosomal contact maps are
distinguished by a single superscript, so we have $\mathbf{C}_{i,i} =
\mathbf{C}_i$.

We denote the number of loci in a chromosome $i$ by $N_i$.
The set of all loci involved in contact map $C_{ij}$ is denoted 
by $\mathbb{V}_{ij}$.
In intra-chromosomal contact maps, $\mathbb{V}_{i,i}$ containts only the 
loci of that particular chromosome ($|\mathbb{V}_i| = N_i$), while in 
inter-chromosomal contact maps $\mathbb{V}_{ij}$ contains the loci in
the both of chromosomes involved 
($\mathbb{V}_{ij} = \mathbb{V}_i \cup \mathbb{V}_j$).

\section{Materials and Methods}
We re-used Leukemic Hi-C libraries 
created in \cite{wang2013properties}
These libraries were sequenced 
for cases of primary human B-acute
lymphoblastic leukemia (B-ALL or ALL), 
the MHH-CALL-4 B-ALL cell
line (CALL4), 
and the follicular lymphoma cell-line (RL).
As in \cite{wang2013properties}, 
We used normal B-cell line (GM068990)
from \cite{lieberman2009comprehensive} 
for our comparisons.
We created contact maps of resolution
500kb and normalized it using 
the \texttt{iced} package in python
developed by \cite{servant2015hic}.
We also performed and extra normalization by
performing network deconvolution \cite{feizi2013network}
in order to remove
the effect of indirect interactions.

@@

Instead, in order to threshold the matrix so that both global and local patterns are
captured, we borrowed the concept of \textit{adaptive thresholding} from image 
processing context. In this method, in order to be set, a pixel should have
an intensity larger than the average of non-zero intensities in its
\textit{neighborhood}. The neighborhood is defined by a sliding kernel 
that passes through the contact map with the pixel at its middle at 
each step. Figure \ref{local_thresholded_chr1_chr1} demonstrates result of 
this thresholding approach for intra-chromosomal contact maps of chromosome 1.
Refer to supplementary material for all 23 interchromosomal thresholding
results.
\begin{figure}[t]
    \centering
    \includegraphics[width=.47\textwidth]{figures/local_thresholded_chr1_chr1.png}
    \caption{Result of thresholding interchromosomal contact map of chromosome 1
    using a kernels of size $5 \times 5$ for all cell lines. 
    The first row shows the thresholded
    maps. Second and third rows demonstrate pair-wise similarities and 
    differences between contact maps respectively.}
    \label{local_thresholded_chr1_chr1}
\end{figure}

\subsection{Orbit Extraction}
Once the thresholded contact maps are obtained, graphlets and orbits can be 
extracted. We used the \texttt{orca} package in \texttt{R} programming 
language to extract the graphlets. As a result of graphlet extraction, 
For each loci in each contact map, a \textit{signature vector} of size
73 is created. Thus for each cell line, we would have 276 
\textit{signature matrices} of
size $|V^{ij}|\times 73$, where $V^{ij}$ is the number of loci
involved in contact map between chromosomes $i$ and $j$. Figure
\ref{fig:graphlet_extraction} illustrates the process and results
of signature matrix extraction schematically.
\begin{figure}
    \centering
    \includegraphics[width=.47\textwidth]{figures/graphlet_extraction.png}
    \caption{Graphlet extraction for the four cell lines. For each
    loci in each contact map between chromosomes $i$ and $j$, 
    the signature vectors of length 73 are extracted, resulting 
    in a \textit{signature matrix} of size $|V^{ij}| \times 73$,
    where $V^{ij}$ 
    is the number of loci involved.}
    \label{fig:graphlet_extraction}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=.5\textwidth]{figures/graphlet_distance_schema.png}
    \caption{Calculating pair-wise loci distances. For each loci (row) 
    in each
    contact map in MIT cell line, its distance is calculated based on
    equation \ref{eq:distance_total} with the corresponding loci in
    leukemic cells. The result of this process is a 
    \textit{signature distance vector} of size
    $|V^{ij}| = N_i+N_j$ for each contact map.
    }
    \label{graphlet_distance_schema}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=.5\textwidth]{figures/graphlet_correlation_schema.png}
    \caption{Calulating pair-wise orbit correlations. For each orbit (column)
    in each contact map in MIT cell line, its correlation with the
    same orbit in the same contact map 
    in leukemic cells is calculated. The result of
    this process is a \textit{signature correlation} vector of size
    73 which captures how similar frequencies of two orbits are.}
    \label{graphlet_correlation_schema}
\end{figure}
%$
%\begin{pmatrix}
%    {S}^A_{\scriptscriptstyle i,j,1,1}         &
%    \cdots                                     &    
%    {S}^A_{\scriptscriptstyle i,j,1,72}        \\
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    \vdots                                     &                
%                                               &      
%    \vdots                                     \\
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   {S}^A_{\scriptscriptstyle i,j,N_1,1}    &
%   \ddots                                            &    
%   {S}^A_{\scriptscriptstyle i,j,N_1, 72} \\
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    \vdots                                     &                
%                                               &      
%    \vdots                                     \\
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   {S}^A_{\scriptscriptstyle i,j,N_1+N_2,1}   &
%    \cdots                                &
%   {S}^A_{\scriptscriptstyle i,j,N_1+N_2, 72}
%\end{pmatrix}
%$
%
%$
%\begin{pmatrix}
%    {C}^A_{\scriptscriptstyle i,j,1,1}     &
%    \cdots    &
%    {C}^A_{\scriptscriptstyle i,j,1,N_1} \\
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    \vdots                             &
%    \ddots&
%    \vdots                             \\
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    {C}^A_{\scriptscriptstyle i,j,N_1,1}   &
%    \cdots&                  
%    S^A_{\scriptscriptstyle i,j,N_1, N_1}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\end{pmatrix}
%$

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{.5\textwidth}
        \includegraphics[width=\textwidth]{figures/orbit-distances_chr14.png}
    \end{subfigure}
    \caption{   
        Pair-wise graphlet signature distances for all contact maps of
        chromosome 14. Warm colors as used for cancer-cancer pairs and
        cold colors are used for normal-cancer pairs. As can be seen
        cancer-cancer pairs tend to be close to each other than to the
        normal cell. This is specially true for the first 13 
        inter-chromosomal contact maps.
     }
    \label{fig:orbit_distances_chr14}
\end{figure}

@@

We performed one-way ANOVA statistical test to 
see if there are significant 
difference between cancer-normal and cancer-cancer pairs.
We found that the difference statistically significant difference.
($F(1, 1654) = 20.49, p < 0.0001$).
As illustrated in figure \ref{fig:orbit_distances_all_a},
we can see that normal-cancer pairs have higher distance
from each other than cancer-cancer pairs.

@@

Figure \ref{fig:orbits_correlations} demonstrates that certain
orbits of Leukemic cells
have higher correlation to each other 
than to the
normal MIT cell. In fact our statistical analysis shows that 
\textit {for orbits NOT in $\mathbb{Q}$, 
intra-leukemic orbit correlations are significantly higher
than leukemic-normal orbit correlations}. This implies
there are significant differences between normal and
leukemic cells in terms of their local structure.
